White Dwarf Mergers on Adaptive Meshes\\ II. Nuclear Reactions

### 2017-09-09 papers/wdmerger_II/paper.tex
Rework the intro using our new theme

Binary white dwarf (WD) systems have long been thought to be potential progenitors
of Type Ia supernovae (SNe Ia) \citep{ibentutukov:1984,webbink:1984}. Recently there
has been a growing amount of observational evidence that double WD systems play
an important role in explaining at least some SNe Ia \citep{maoz:2014}. These systems
may be isolated or they may be in hierarchical triple systems, where a WD binary
system in a tight inner orbit is gravitationally coupled to a third, outer star.
In either case the binary WD pair can merge due to the emissions of gravitational waves,
but in the hierarchical triple case, dynamical interactions between the WD binary and
the outer star can prompt mergers as well \citep{thompson:2011,hamers:2013}. The mergers
occur due to perturbations in the eccentricity of the binary orbit caused by the outer
star, which significantly increase the gravitational wave emission of the system. In a
subset of these mergers the eccentricity may be driven to such a high value that the
merger resembles a head-on collision of the binary WDs. Whether the rate of these collisions
may be high enough to significantly contribute to the observed SN Ia rate is debated:
\cite{katzdong:2012} suggest that this is plausible, but \cite{hamers:2013} and
\cite{papish:2015} suggest that the rate is relatively low. Even if the rate is low,
if at least some SNe Ia are caused by WD collisions then they are worth understanding.

Over the past few years a number of groups have performed simulations of WD collisions
to understand whether they may yield astrophysical transients that look like SNe Ia
\citep{rosswog:2009,raskin:2010,loren-aguilar:2010,hawley:2012,garcia-senz:2013,
kushnir:2013,papish:2015,holcomb:2015}. Head-on collisions rapidly convert a significant
amount of kinetic energy into thermal energy in a small region and thus set up the
conditions for a thermonuclear detonation, and these previous simulations have indeed
found that detonations occur and convert a large amount of carbon/oxygen material into
iron-group elements. These papers varied significantly in the hydrodynamic methods used
(Lagrangian versus Eulerian methods), the methodology used for coupling nuclear
reactions to the hydrodynamics (including variation in the number of isotopes in
the nuclear network and in the evolution of the temperature), and the temporal and
spatial resolution. Consequently they yielded estimates of production of nickel-group
elements that varied significantly, leaving much uncertainty about how such an event
would appear observationally and whether it bears any resemblance to a SN Ia.
See Table 4 of \cite{garcia-senz:2013} for a summary of the outcome of many of these studies.

In this paper we seek to add another view to the problem using the reactive hydrodynamics
code \castro\ \citep{castro} and its underlying adaptive mesh refinement library \amrex\footnote{\amrex is the successor to \boxlib.}.
This paper is a follow-up to \citet{wdmergerI} (hereafter, Paper I), which established
the hydrodynamic methdology we are using to simulate mergers of white dwarfs.
Paper I left out a discussion of the methodology we are using for nuclear reactions,
and we fill out this information below in \autoref{sec:numericalmethods}. While it is
our aim is to add information to the discussion about collisions of white dwarfs,
we also view this work as a methodology study that asks to what extent we
can trust the reliability of explosive thermonuclear reactions in Eulerian
hydrodynamics simulations. As a corollary, to what extent does static and/or
adaptive mesh refinement resolve the types of problems that occur in low resolution
simulations? We discuss this in Section \autoref{sec:resolution}, though first
we examine the response of the code to a number of user-specified parameters in
\autoref{sec:parameters}. We then close with our thoughts on what our work implies
for the broader astrophysical reacting hydrodynamics literature in \autoref{sec:conclusion}.

@@

The hydrodynamical equations solved in this paper were presented in \citet{wdmergerI} (hereafter, Paper I).
Revisions to the core algorithm since then are presented in \autoref{app:hydro}.
In this paper we focus our discussion on our implementation of the nuclear
reaction network and integrator. There are three main areas of concern: first,
specifying what nuclides are in the network and what reaction rates link the
various nuclides; second, once the system of ODEs governing the evolution of
the nuclear species has been written down, how to couple them to ODEs governing
the evolution of the thermodynamics, in particular the temperature; third,
coupling the nuclear reaction updates to the evolution of the hydrodynamical
system. We will deal with each of these in turn in the following subsections.

@@

We implement the white dwarf collision problem in \castro\ using our \wdmerger\
package. We load the grid in the way most previous papers have done so:
the white dwarf centers of mass are initially separated by a distance of four times
the (secondary) white dwarf radius. Note that the radius of the white dwarf, and
consequently the initial distance, depends on the equation of state used, and this
can vary somewhat depending on the relevant physics used, such as Coulomb corrections
in the Helmholtz equation of state. However, the results do not really depend on the
exact initial distance as long as there is enough time for the WDs to distort in
response to tidal forces as they approach. Their initial velocity is that of
two point-masses in free-fall towards each other coming in from infinity, such that
the contact point is at the origin, and they approach each other along the $x$-axis.
The composition of the WDs is as described in Paper I: low-mass WDs are helium-dominated,
intermediate-mass WDs are carbon and oxygen dominated, and high-mass WDs contain
oxygen, neon, and magnesium. By default the WDs approach each other head-on, with
their respective centers lying on the $x$-axis, but we have also added an option
to have the WDs approach each other at a non-zero impact parameter $b$, where the
offset is perpendicular to the $x$-axis in the $y$ direction. We measure
$b$ in units of the secondary WD's radius, so that the default head-on case
corresponds to $b = 0$ and the case where the two WDs just graze past each other
corresponds to $b = 1$ (in practice it happens slightly differently because
of the WDs responding to each other's gravitational fields).

>>>>

Binary white dwarf (WD) systems are a promising progenitor candidate for Type
Ia supernovae (SNe Ia) for observational reasons, but from the simulation
perspective there is still significant uncertainty regarding how nuclear
detonations form in an exploding white dwarf. There is uncertainty in the
model, regarding whether detonations form on the surface of a star as
accreting material heats up when it strikes the surface, or at or near the
center of the star as the accreted material compresses the WD. There is also
uncertainty in the numerics: given one of these models, can we self-consistently
demonstrate that a detonation occurs? This question has haunted both single
WD and double WD progenitor models, because the length and time scale at which
a detonation forms is orders of magnitude smaller than the resolution that
typical hydrodynamic simulations can achieve. Furthemore, the approximate equations
we use at the resolution that we can achieve are still notoriously
difficult to solve accurately. The mere presence of a detonation
(or lack thereof) in a simulation is therefore only weak evidence regarding
whether a detonation would truly occur.

In this study we examine the challenges associated with simulating nuclear detonations.
This paper follows from our first WD merger methodology paper, \citet{wdmergerI}
(hereafter, Paper I), where we deferred discussion of our nuclear reaction methodology.
After presenting the burning methodology, we apply it to the problem of collisions of white dwarfs.
WD collisions can occur in isolated binaries or in hierarchical triple systems, where the WD binary
system in a tight inner orbit is gravitationally coupled to a third, outer star \citep{thompson:2011,hamers:2013}.
Whether the rate of these collisions may be high enough to significantly contribute to
the observed SN Ia rate is debated: \cite{katzdong:2012} suggest that this is plausible,
but \cite{hamers:2013} and \cite{papish:2015} suggest that the rate is relatively low.
Regardless, they are a useful vehicle for understanding the problem of nuclear burning.
Head-on WD collisions rapidly convert a significant amount of kinetic energy into thermal
energy in a small region and thus set up conditions ripe for a thermonuclear detonation;
being able to correctly simulate these is in some sense a prerequisite to correct simulation
of the more complicated general case of white dwarf mergers, where we do not initially know
under what conditions the seed of a thermonuclear detonation is born.

There have been a number of previous studies which provide useful insight into setting up
the problem and the burning methodology \citep{rosswog:2009,raskin:2010,loren-aguilar:2010,
hawley:2012,garcia-senz:2013,kushnir:2013,papish:2015,holcomb:2015}. These studies
found that detonations occur and convert a large amount of carbon/oxygen material into
iron-group elements. These papers varied significantly in the hydrodynamic methods used
(Lagrangian versus Eulerian methods), the methodology used for coupling nuclear
reactions to the hydrodynamics (including variation in the number of isotopes in
the nuclear network and in the evolution of the temperature), and the temporal and
spatial resolution. Consequently the detonations they formed varied in location and nature,
and the resultant estimates of production of nickel-group elements varied significantly,
leaving much uncertainty about how such an event would appear observationally and whether
it bears any resemblance to a SN Ia. See Table 4 of \cite{garcia-senz:2013} for a summary
of the outcome of many of these studies. In this paper we make the case that
while we can identify the main problems prevent accurate simulation of the collision
event, there is no clear path to overcoming them using the computational resources
typically available for contemporary hydrodynamics simulations.

@@

The hydrodynamical equations solved in this paper were presented in \citet{wdmergerI}
(hereafter, Paper I) for the reacting hydrodynamics code \castro\ \citep{castro}.
Revisions to the core algorithm since then are presented in \autoref{app:hydro}.
In this paper we focus our discussion on our implementation of the nuclear
reaction network and integrator. There are three main areas of concern: first,
specifying what nuclides are in the network and what reaction rates link the
various nuclides; second, once the system of ODEs governing the evolution of
the nuclear species has been written down, how to couple them to ODEs governing
the evolution of the thermodynamics, in particular the temperature; third,
coupling the nuclear reaction updates to the evolution of the hydrodynamical
system. We will deal with each of these in turn in the following subsections.

@@

We implement the white dwarf collision problem in \castro\ in the \wdmerger\ problem
directory. We prepare the domain in the way most previous papers have done so:
the white dwarf centers of mass are initially separated by a distance of four times
the (secondary) white dwarf radius. Note that the radius of the white dwarf, and
consequently the initial distance, depends on the equation of state used, and this
can vary somewhat depending on the relevant physics used, such as Coulomb corrections
in the Helmholtz equation of state. However, the results do not really depend on the
exact initial distance as long as there is enough time for the WDs to distort in
response to tidal forces as they approach. Their initial velocity is that of
two point-masses in free-fall towards each other coming in from infinity, such that
the contact point is at the origin, and they approach each other along the $x$-axis.
The composition of the WDs is as described in Paper I: low-mass WDs are helium-dominated,
intermediate-mass WDs are carbon and oxygen dominated, and high-mass WDs contain
oxygen, neon, and magnesium. By default the WDs approach each other head-on, with
their respective centers lying on the $x$-axis, but we have also added an option
to have the WDs approach each other at a non-zero impact parameter $b$, where the
offset is perpendicular to the $x$-axis in the $y$ direction. We measure
$b$ in units of the secondary WD's radius, so that the default head-on case
corresponds to $b = 0$ and the case where the two WDs just graze past each other
corresponds to $b = 1$ (in practice it happens slightly differently because
of the WDs responding to each other's gravitational fields).
